---
title: "Getting Started with lehdr"
subtitle: "Accessing and Analyzing LODES Data in R"
author:
  - name: Jamaal Green
  - name: Dillon Mahmoudi
  - name: Liming Wang
package: lehdr
date: "`r format(Sys.time(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lehdr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Introduction

**lehdr** is an R package that provides programmatic access to the
[Longitudinal Employer-Household Dynamics (LEHD)](https://lehd.ces.census.gov/data/#lodes)
Origin-Destination Employment Statistics (LODES) dataset. LODES forms the
backbone of the U.S. Census Bureau's
[OnTheMap](https://OnTheMap.ces.census.gov/) web application, which maps
commuting patterns at fine geographic scales. While OnTheMap is a useful
exploratory tool, it does not support systematic comparisons over time or
across geographies. **lehdr** fills that gap by exposing LODES downloads as
tidy data frames and providing built-in functions for common analyses.

LODES contains three file types:

- **OD** (`lodes_type = "od"`): Origin-destination flows, giving the number
  of workers by home-workplace Census block pair.
- **RAC** (`lodes_type = "rac"`): Residential area characteristics, giving
  job totals at workers' home blocks.
- **WAC** (`lodes_type = "wac"`): Workplace area characteristics, giving
  job totals at workers' job blocks.

## Installation

```{r install, eval=FALSE}
# From CRAN
install.packages("lehdr")

# Development version from GitHub
devtools::install_github("jamgreen/lehdr")

library(lehdr)
library(dplyr)
```

## Basic Usage

### Downloading a single state and year

The following example retrieves 2020 Oregon OD data aggregated to the Census
tract level. Only primary jobs (`job_type = "JT01"`) across all worker
segments (`segment = "S000"`) are included.

```{r usage_basic}
or_od <- grab_lodes(
  state      = "or",
  year       = 2020,
  lodes_type = "od",
  job_type   = "JT01",
  segment    = "S000",
  state_part = "main",
  agg_geo    = "tract"
)

head(or_od)
```

### Multiple states and years

Pass character vectors to `state` and/or numeric vectors to `year`. The
function returns a single bound tibble across all combinations.

```{r usage_multi}
or_ri_od <- grab_lodes(
  state      = c("or", "ri"),
  year       = c(2019, 2020),
  lodes_type = "od",
  job_type   = "JT01",
  segment    = "S000",
  state_part = "main",
  agg_geo    = "tract"
)

head(or_ri_od)
```

Not all state-year combinations are available for every LODES version. Consult
the [LEHD Technical Document](https://lehd.ces.census.gov/data/lodes/LODES8/)
for availability.

### Aggregation levels

`agg_geo` controls the output geography. Valid values are `"block"` (default),
`"bg"` (block group), `"tract"`, `"county"`, and `"state"`. Aggregation
truncates GEOID strings and sums all numeric columns.

```{r usage_agg}
md_rac_county <- grab_lodes(
  state      = "md",
  year       = 2019,
  lodes_type = "rac",
  job_type   = "JT01",
  segment    = "S000",
  agg_geo    = "county"
)

head(md_rac_county)
```

### Attaching spatial geometries

Set `geometry = TRUE` to retrieve Census geometries via the `tigris` package
and attach them to the result. For `lodes_type = "rac"` or `"wac"`, the
return value is an `sf` object. For `lodes_type = "od"`, two geometry columns
(`h_geometry`, `w_geometry`) are added but the object remains a tibble ---
use `sf::st_as_sf()` to promote it.

```{r usage_geometry}
md_wac_sf <- grab_lodes(
  state      = "md",
  year       = 2019,
  lodes_type = "wac",
  job_type   = "JT01",
  segment    = "S000",
  agg_geo    = "county",
  geometry   = TRUE
)

plot(md_wac_sf["C000"])
```

### Geographic crosswalk

`grab_crosswalk()` downloads the LODES geographic crosswalk, which maps
Census blocks to all higher-level geographies. Useful when you need custom
aggregations not covered by `agg_geo`.

```{r usage_xwalk}
vt_xwalk <- grab_crosswalk("vt")

# LODES7 vintage crosswalk (2010 Census blocks)
vt_xwalk_7 <- grab_crosswalk("vt", version = "LODES7")
```

---

## Analytical Functions

### Commute flow statistics (Option A)

`compute_commute_stats()` summarizes commute flows from an OD dataset into
four per-geography metrics: inflow, outflow, internal flows, net flow, and
the self-containment ratio. Self-containment is the share of employed
residents who also work in the same geographic unit.

```{r commute_stats}
od_md <- grab_lodes(
  state      = "md",
  year       = 2019,
  lodes_type = "od",
  job_type   = "JT00",
  segment    = "S000",
  state_part = "main",
  agg_geo    = "county"
)

commute_md <- compute_commute_stats(od_md, agg_geo = "county")

# Most self-contained counties
commute_md %>%
  arrange(desc(self_containment)) %>%
  select(county, workers_in, workers_out, net_flow, self_containment) %>%
  head(10)
```

Net flow identifies counties that are net job importers (positive) versus
net exporters (negative), a key indicator of job/housing imbalance.

### Longitudinal change (Option B)

`compute_lodes_change()` computes absolute and percentage change in any LODES
variable between two years. The output can be wide (one row per geography,
columns for each variable's change) or long (one row per geography-variable
combination, suitable for `ggplot2`).

```{r lodes_change}
wac_multi <- grab_lodes(
  state      = "md",
  year       = c(2010, 2019),
  lodes_type = "wac",
  job_type   = "JT00",
  segment    = "S000",
  agg_geo    = "county"
)

change_md <- compute_lodes_change(
  wac_multi,
  geo_col      = "w_county",
  base_year    = 2010,
  compare_year = 2019,
  variables    = c("C000", "CE01", "CE02", "CE03")
)

# Counties with the largest absolute job growth
change_md %>%
  arrange(desc(C000_change)) %>%
  select(w_county, C000_base, C000_compare, C000_change, C000_pct_change) %>%
  head(10)
```

Long format is useful for visualizing change across multiple variables:

```{r lodes_change_long, eval=FALSE}
library(ggplot2)

change_long <- compute_lodes_change(
  wac_multi,
  geo_col  = "w_county",
  output   = "long",
  variables = c("CE01", "CE02", "CE03")
)

ggplot(change_long, aes(x = w_county, y = pct_change, fill = variable)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Percent change in jobs by earnings tier, Maryland counties, 2010-2019",
    x = NULL, y = "Percent change"
  )
```

### Earnings tier shares (Option D)

`compute_earnings_share()` computes the distribution of jobs (or resident
workers) across three monthly earnings tiers: low (up to $1,250/month),
mid ($1,251-$3,333/month), and high (above $3,333/month). These tiers
correspond to the `CE01`/`CE02`/`CE03` columns in WAC data and
`CR01`/`CR02`/`CR03` in RAC data.

```{r earnings_share}
wac_md <- grab_lodes(
  state      = "md",
  year       = 2019,
  lodes_type = "wac",
  job_type   = "JT00",
  segment    = "S000",
  agg_geo    = "county"
)

earn_shares <- compute_earnings_share(wac_md, type = "wac", geo_col = "w_county")

# Counties with highest share of low-wage jobs
earn_shares %>%
  arrange(desc(share_low)) %>%
  select(w_county, share_low, share_mid, share_high) %>%
  head(10)
```

Combining `compute_earnings_share()` and `compute_lodes_change()` allows
longitudinal tracking of wage polarization:

```{r earnings_share_change, eval=FALSE}
wac_panel <- grab_lodes(
  state      = "md",
  year       = c(2010, 2019),
  lodes_type = "wac",
  job_type   = "JT00",
  segment    = "S000",
  agg_geo    = "county"
)

# Compute shares for each year separately, then use compute_lodes_change
# on the resulting share columns for a custom polarization analysis.
share_2010 <- compute_earnings_share(
  dplyr::filter(wac_panel, year == 2010),
  type = "wac", geo_col = "w_county"
)

share_2019 <- compute_earnings_share(
  dplyr::filter(wac_panel, year == 2019),
  type = "wac", geo_col = "w_county"
)
```

---

## Advanced: Job Accessibility Index (Illustrative)

The following is an illustrative example of a simple gravity-based job
accessibility index computed entirely from LODES OD data. It is included
here as a starting point; a full accessibility analysis typically requires
additional data on travel times or distances (see packages such as `r5r` or
`accessibility`).

The simplest accessibility measure is a raw count: for a given home
geography, the number of reachable jobs. Using OD data, you can approximate
a threshold-based accessibility score by counting the number of distinct
work geographies connected to each home geography (i.e., the number of
destinations with non-zero flows), weighted by flow volume.

```{r accessibility, eval=FALSE}
od_md <- grab_lodes(
  state      = "md",
  year       = 2019,
  lodes_type = "od",
  job_type   = "JT00",
  segment    = "S000",
  state_part = "main",
  agg_geo    = "tract"
)

# Simple gravity-style accessibility: for each home tract, sum jobs
# reachable (weighted by S000 flow counts, as a proxy for connectivity)
accessibility <- od_md %>%
  dplyr::group_by(h_tract) %>%
  dplyr::summarise(
    n_dest         = dplyr::n_distinct(w_tract),
    total_reachable = sum(S000, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(total_reachable))

head(accessibility)
```

`n_dest` captures the breadth of access (number of distinct job destinations
a resident tract sends workers to), while `total_reachable` is a
flow-weighted measure of aggregate connectivity. These can be mapped using
`tigris` geometries via `grab_lodes(..., geometry = TRUE)`.

---

## Caching

By default, downloaded files are deleted after reading. Setting
`use_cache = TRUE` retains files in the user-level cache directory
(`tools::R_user_dir("lehdr", "cache")`), which avoids re-downloading when
you call `grab_lodes()` repeatedly with the same parameters. You can also
set this option globally for a session:

```{r caching}
options(lehdr_use_cache = TRUE)
```
